\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-26 paper/examples.tex

The epidemiologist is excited
to get a quick and accurate view into the data. However,
the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets}),
and there is a significant amount of missing data across all
relevant attributes.

Before she can perform her queries, the researcher must develop a strategy
for handling the missing values. She currently has two options:
1) she could drop records with missing values in relevant fields,
2) she could use a standard data imputation package on all of her data. Both of these
approaches have significant drawbacks. For the query pictured in 
Fig.~\ref{fig:example-query}, (1) drops 1492 potentially relevant tuples,
while her previous experiences with (2) have proven be expensive. In a previous project, she
worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}. The analyst needs a more complete picture
of the data, so (1) is insufficient, and she is under time pressure, so (2)
is infeasible.

She can run her queries immediately and finish her report if
she uses \ProjectName{}, which allows her to write her
query in standard SQL and automatically performs the imputations necessary to fill in the missing data.
An example query is shown in Fig.~\ref{fig:example-query}.

@@

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\vspace{0.5\baselineskip}
\caption{A quality-optimized plan for the query in Fig.~\ref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
\label{fig:quality-plan}
\end{figure}

@@

For the query in Fig.~\ref{fig:example-query}, \ProjectName{} generates
the plans in Fig.~\ref{fig:quality-plan} and~\ref{fig:fast-plan}, when 
optimizing for quality and performance, respectively.

The quality-optimized plan (Fig.~\ref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, the performance-optimized plan (Fig.~\ref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\caption{A performance-optimized plan for the query in Fig.~\ref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
\label{fig:fast-plan}
\end{figure}

>>>>

The epidemiologist is excited
to get a quick and accurate view into the data. However,
the data has been collected through CDC surveys (see~\Cref{subsec:datasets}),
and there is a significant amount of missing data across all
relevant attributes.

Before she can perform her queries, the researcher must develop a strategy
for handling the missing values. She currently has two options:
1) she could drop records with missing values in relevant fields,
2) she could use a standard data imputation package on all of her data. Both of these
approaches have significant drawbacks. For the query pictured in ~\Cref{fig:example-query}, 
(1) drops 1492 potentially relevant tuples,
while her previous experiences with (2) have proven be expensive. In a previous project, she
worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}. The analyst needs a more complete picture
of the data, so (1) is insufficient, and she is under time pressure, so (2)
is infeasible.

She can run her queries immediately and finish her report if
she uses \ProjectName{}, which allows her to write her
query in standard SQL and automatically performs the imputations necessary to fill in the missing data.
An example query is shown in~\Cref{fig:example-query}.

@@

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\vspace{0.5\baselineskip}
\caption{A quality-optimized plan for the query in~\Cref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
\label{fig:quality-plan}
\end{figure}

@@

For the query in~\Cref{fig:example-query}, \ProjectName{} generates
the plans in~\Cref{fig:quality-plan} and~\Cref{fig:fast-plan}, when 
optimizing for quality and performance, respectively.

The quality-optimized plan (\Cref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, the performance-optimized plan (\Cref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
\label{fig:fast-plan}
\end{figure}

### 2017-02-26 paper/introduction.tex

\todobox{see how this bit fits}{
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.}

% Prior work
There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
Prior work in statistics has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
\ProjectName{} contributes a novel approach to planning data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
and express preferences in quality of data imputation versus running time.

>>>>

% Prior work
There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
Prior work in statistics has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
\ProjectName{} contributes a novel approach to planning data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
and express preferences in quality of data imputation versus running time. The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
